---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r prefix_chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# oefar

<!-- badges: start -->
[![R-CMD-check](https://img.shields.io/badge/R--CMD--check-passing-brightgreen.svg)](https://github.com/)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-version](https://img.shields.io/badge/R-%3E%3D%204.1.0-blue.svg)](https://www.r-project.org/)
[![R-CMD-check](https://github.com/PaulESantos/oefar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/oefar/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PaulESantos/oefar/graph/badge.svg)](https://app.codecov.io/gh/PaulESantos/oefar)
<!-- badges: end -->

**`oefar`** is an R package designed to facilitate access, exploration, and direct downloading of open data from the **Agency for Environmental Assessment and Enforcement (OEFA)** of Peru via its v2 Open Data API ([datosabiertos.oefa.gob.pe](https://datosabiertos.oefa.gob.pe/)).

The package provides **98 dedicated functions** for each dataset in the official catalog, automatic column name cleaning into snake_case according to tidyverse conventions, configurable timeout handling, and native `tibble` support.

---

## Installation

You can install the released version of `oefar` from CRAN with:

```r
install.packages("oefar")
```

Or install the development version from GitHub using `remotes` or `devtools`:

```r
# install.packages("remotes")
remotes::install_github("PaulESantos/oefar")
```

---

## API Key Configuration

To use the OEFA Open Data API, a **personal API Key** is required. You can obtain one directly from the OEFA developer portal: [https://datosabiertos.oefa.gob.pe/developers/](https://datosabiertos.oefa.gob.pe/developers/).

### Temporary configuration for current session:
```r
library(oefar)

# Set the API Key for the current R session
oefa_set_api_key("YOUR_API_KEY_HERE")
```

### Permanent configuration in local R environment (`.Renviron`):
It is recommended to add your key to your `.Renviron` file as `OEFA_API_KEY`:

```env
OEFA_API_KEY="YOUR_API_KEY_HERE"
```

The package will automatically detect the key via `oefa_get_api_key()`.

---

## Usage Examples

### 1. Catalog Exploration

```{r}
library(oefar)

# List all OEFA data categories
categories <- oefa_list_categories()
print(categories)

# List all available datasets (datastreams)
catalog <- oefa_list_datastreams()
head(catalog[, c("guid", "title", "category_name")])

# Search datasets by keyword (e.g., "denuncias", "agua", "supervision")
complaints_ds <- oefa_search_datastreams("denuncias")
print(complaints_ds[, c("guid", "title")])
```

### 2. Downloading Datasets with Dedicated Functions

The package includes **98 dedicated helper functions** prefixed with `oefa_get_<dataset_name>()` for direct download:

```{r}
# Download SINADA complaints dataset
complaints <- oefa_get_denun_sinad(limit = 100)
print(complaints)

# Download Personnel Selection Processes
personnel <- oefa_get_proce_de_selec_de_71611(limit = 50)
print(personnel)

# Download Environmental Monitoring - Water (EAC)
water <- oefa_get_eac_compo_ambie_agua(limit = 100)
print(water)
```

### 3. Adjusting Request Timeout

For large data downloads or higher-latency connections, you can adjust the request timeout parameter:

```{r}
# Extend wait timeout to 120 seconds for a large dataset
large_complaints <- oefa_get_denun_sinad_61293(timeout = 120)

# Or using the generic oefa_get_data() function
generic_data <- oefa_get_data(guid = "DENUN-SINAD-61293", timeout = 180)
```

---

## Data Attribution and Official Reference

Data distributed and consumed through this package belongs to and is publicly provided by the **Agency for Environmental Assessment and Enforcement (OEFA)** of the Government of Peru.

### Citing the Package and Data Source

In R, you can view formal citation details by running:

```{r, eval = FALSE}
citation("oefar")
```

#### R Package Citation:
> Santos Andrade, P. E. (2026). *oefar: Access to the 'OEFA' Open Data API of Peru*. R package version 0.1.0. URL: https://datosabiertos.oefa.gob.pe/.

#### Official Data Source Reference (OEFA):
> Agency for Environmental Assessment and Enforcement - OEFA. (2026). *OEFA Open Data Portal* [Online database]. Government of Peru. URL: https://datosabiertos.oefa.gob.pe/.

---

## License

This package is licensed under the [MIT License](https://opensource.org/licenses/MIT). Environmental enforcement data is property of the Peruvian state (OEFA) and is subject to the terms of Peru's National Open Data Platform.
